Fast gravitational wave parameter estimation without compromises

### 2023-08-01 src/tex/ms.tex
Slightly editing corrections for clarity

% More sample helps training normalizing flow
Yet, in our work, the use of accelerators provides two independent advantages
that tremendously benefit the parameter estimation process. First, using
accelerators allows us to run many independent MCMC chains simultaneously, which
benefits the training of the normalizing flow. Since we generate the data we use
to train the normalizing flow on the fly, the more independent data we can feed
to the training process, the higher chance the normalizing flow can learn a
reasonable representation of the target distribution.  If we only used a small
number of chains, we would be limited to the correlated samples from each chain
and we would have to run more sequential steps to obtain the same amount of
independent samples in the end---with more chains the problem becomes
parallelizable and we can obtain the same number of training samples sooner. In
other words, \corr{being able to use many independent chains helps the normalizing
flow approximating the target density faster in wall time.}

@@

\corr{The interplay between \textsc{flowMC} and GPU acceleration is critical to
the performance of the code. A standard PE run using a nested sampler or
standard MCMC requires around $10^6$ waveform evaluations, Assuming $10^5$
frequency bins per waveform, that corresponds to evaluating the waveform model
at $10^11$ frequency-parameter points. One would think this can be achieved
within $\sim 100\textrm{s}$ by a A100 GPU given the benchmark mentioned above.
However, a sampler requires sequential evaluation to accept or reject a sample,
where what a GPU provide is parallel evaluation of the waveform, i.e. it can run
many chains in parallel, but not many steps with the same amount of computation.
This means one cannot just obtain the posterior samples by running the waveform
model on a GPU with a typical sampler, in the sense that we cannot run many
chains for one step and expect unbias posterior samples.
This is precisely why \textsc{flowMC} is necessary for our approach. Since every
accepted sample generated using the normalizing flow as proposal is nearly independent,
and the sampling from the normalizing flow is an embarrassingly parallel
process, the name of the game then becomes training the normalizing flow as fast as
possible. Having many chains helps to produce the necessary amount of samples
for training the normalizing flow, which means now we can meaningfully harvest
the benefit of having large throughput of GPU. Without either of this two,
our approach would be significantly less performant.}

@@

To demonstrate the performance of our parameter estimation pipeline, we apply it
to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors shown in
Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz starting at
20 Hz for GW150914, and 128 s of data sampled at 4096 Hz starting at 23 Hz for
GW170817; strain data and PSDs for both events are fetched from GWOSC
\citep{GWOSC}. We use the same normalizing flow model as the injection-recovery
study. For our specific choice of sampler settings, we produce ${\sim}$2500 and
3500 \emph{effective samples}\footnote{Effective samples here refers to the
number of independent samples, which is the total number of generated samples
divided by their correlation length; we compute the effective sample size using
\textsc{arviz} \citep{arviz_2019},
\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}. } for
GW150914 and GW170817 respectively. \corr{Running on a node with an
\textsc{Nvidia} 80GB A100 GPU and 16 intel skylake CPUs} \footnote{Note that the
CPU time is hard to benchmark in \textsc{Jax} due to JIT compilation and
asynchronous dispatch. Overall, we find 4 CPUs to be sufficient for our use
case. We chose 16 because of the ease of use of the local cluster.}, the wall
time for both events is around 3 minutes. Most of this time is spent on
just-in-time (JIT) compilation of the code; the actual sampling time is only
${\sim}40$ s. We pre-compute the reference waveform parameters used to
hetetrodyne the likelihood for the two events, which is omitted in the wall-time
calculation. \corr{In terms of specific configuration of the sampler, we used
1000 chains for both of the events. We run both the tuning phase and the
production phase for 20 loops each. Within each phase, the global sampler and
local sampler are stepped 200 times. This translates into a total number of
likelihood evaluation is $1.6*10^7$. \footnote{While in principle adding more
chains makes the algorithm less efficient in terms of acceptance/evaluation
ratio, since GPUs can evaluate a fixed number of instruction per cycle, we can
keep adding chains without paying extra computational cost before we saturate
the GPU. This results in an improvement in wall time while being less efficient.
We decide to choose a setting that improve wall time instead of efficiency.} The
acceptance rate of the global proposal is about $2\%$ and $5\%$ for GW150914 and
GW170817 respectively. Note that the setting is chosen to run longer than
necessary to ensure convergence, we are currently pursing tuning studies to
understand trade-offs in the algorithms and improve the efficiency further.}

@@

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files available on github
\url{https://github.com/kazewong/TurboPE/tree/main/src}.  We carry out these
runs using \textsc{parallel Bilby} (\textsc{pBilby}) \citep{Smith:2019ucc} to
distribute the computation over 400 Intel Skylake CPUs for each event. For the
specific settings chosen, the wall-time duration of each run was ${\sim}2$ h for
GW150914 and ${\sim}1$ day for GW170817. \corr{We have also benchmarked the
performance using a ROQ likelihood, which takes about 30 minutes for GW170817.
Without ROQ, it takes about a week to complete a run. This broadly agrees with
what is suggested in the literature \cite{Smith:2016qas}}

@@

% Additional traits for having differentiable waveforms
We are currently working on a number of improvements and extensions to our
current infrastructure. While the \textsc{IMRPhenomD} waveform approximant is a
reasonable start, it lacks some qualitative features that other state-of-the-art
models have, such as precession, subdominants moments of the radiation, tides,
or eccentricity. It also has a higher mismatch with reference numerical
relativity waveforms compared to more recent waveform models. Currently, we are
working on building differentiable implementations of \textsc{IMRPhenomXPHM}
\citep{Pratten:2020ceb}, a precessing successor to \textsc{IMRPhenomD} including
subdominant harmonics, as well as the numerical relativity surrogate waveforms,
including \textsc{NRSur7dq4} \citep{Varma:2018mmi}. Going forward, we expect the
use of autodifferentiation environments like \textsc{Jax} to become more
prevalent in the waveform development community, increasing the number of
differentiable waveform models available. This would not only be beneficial for
parameter estimation, but also for a number of other applications such as Fisher
matrix computations, template placement and calibrating waveforms to numerical
relativity
\citep{Coogan:2022qxs,Iacovelli:2022bbs,Iacovelli:2022mbg,Edwards:2023sak}.
\corr{Another family of waveform models that could benefit from this approach is
the effective-one-body (EOB) models \citep{Buonanno:1998gg,PhysRevD.89.061502}.
However, the EOB family requires solving time-dependent ODEs to compute the
waveform and the truncation of the ODEs is dynamically determined at runtime.
JIT in \textsc{Jax} requires the computational graph of a function to be fixed
at compile time, which means the ODEs solving part of EOB is not compatible with
\textsc{Jax}, so the models in the EOB family may not benefit as much as the
other families.}

>>>>

% More sample helps training normalizing flow
Yet, in our work, the use of accelerators provides two independent advantages
that tremendously benefit the parameter estimation process. First, using
accelerators allows us to run many independent MCMC chains simultaneously, which
benefits the training of the normalizing flow. Since we generate the data we use
to train the normalizing flow on the fly, the more independent data we can feed
to the training process, the higher chance the normalizing flow can learn a
reasonable representation of the target distribution.  If we only used a small
number of chains, we would be limited to the correlated samples from each chain
and we would have to run more sequential steps to obtain the same amount of
independent samples in the end---with more chains the problem becomes
parallelizable and we can obtain the same number of training samples sooner. In
other words, \corr{being able to use many independent chains helps the normalizing
flow to approximate the target density in a shorter wall time.}

@@

\corr{The interplay between \textsc{flowMC} and GPU acceleration is critical to
the performance of the code. A standard PE run using a nested sampler or
standard MCMC requires around $10^6$ waveform evaluations. Assuming $10^5$
frequency bins per waveform, this corresponds to evaluating the waveform model
at $10^{11}$ frequency-parameter points. One would think this can be achieved
within $\sim 100\textrm{s}$ by an A100 GPU given the benchmark mentioned above.
However, typical samplers require sequential evaluations to accept or reject samples,
where what a GPU provides is parallel evaluation of the waveform i.e., in a fixed amount of time it can run
many chains in parallel, but not many sequential steps.
This means one cannot obtain the posterior samples by running the waveform
model on a GPU with a typical sampler. For example, in the most extreme case this would mean running many
chains for just one step which would obviously lead to biased posterior samples.
This is precisely why \textsc{flowMC} is a key ingredient in our approach. Since every
accepted sample generated using the normalizing flow as the proposal is nearly independent,
and the sampling from the normalizing flow is an embarrassingly parallel
process, the name of the game then becomes training the normalizing flow as fast as
possible. Having many chains helps to produce the necessary amount of samples
for training the normalizing flow, which means we can meaningfully harvest
the benefit of having the large throughput of the GPU.}

@@

To demonstrate the performance of our parameter estimation pipeline, we apply it
to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors shown in
Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz starting at
20 Hz for GW150914, and 128 s of data sampled at 4096 Hz starting at 23 Hz for
GW170817; strain data and PSDs for both events are fetched from GWOSC
\citep{GWOSC}. We use the same normalizing flow model as the injection-recovery
study. For our specific choice of sampler settings, we produce ${\sim}$2500 and
3500 \emph{effective samples}\footnote{Effective samples here refers to the
number of independent samples, which is the total number of generated samples
divided by their correlation length; we compute the effective sample size using
\textsc{arviz} \citep{arviz_2019},
\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}. } for
GW150914 and GW170817 respectively. \corr{Running on a node with an
\textsc{Nvidia} 80GB A100 GPU and 16 intel skylake CPUs} \footnote{Note that the
CPU time is hard to benchmark in \textsc{Jax} due to JIT compilation and
asynchronous dispatch. Overall, we find 4 CPUs to be sufficient for our use
case. We chose 16 because of the ease of use of the local cluster.}, the wall
time for both events is around 3 minutes. Most of this time is spent on
just-in-time (JIT) compilation of the code; the actual sampling time is only
${\sim}40$ s. We pre-compute the reference waveform parameters used to
hetetrodyne the likelihood for the two events, which is omitted in the wall-time
calculation. \corr{In terms of specific configuration of the sampler, we used
1000 chains for both of the events. We run both the tuning phase and the
production phase for 20 loops each. Within each phase, the global sampler and
local sampler are stepped 200 times. This translates into $1.6 \times 10^7$
likelihood evaluations in total.\footnote{While in principle adding more
chains makes the algorithm less efficient in terms of the acceptance/evaluation
ratio, since GPUs can evaluate a fixed number of instructions per cycle, we can
keep adding chains without paying extra computational cost before we saturate
the GPU. This results in an improvement in wall time while being less efficient.
We decide to choose a setting that improves the wall time instead of efficiency.} The
acceptance rate of the global proposal is about $2\%$ and $5\%$ for GW150914 and
GW170817 respectively. Note that the setting is chosen to run longer than
necessary to ensure convergence; we are currently pursing tuning studies to
understand trade-offs in the algorithms and improve the efficiency further.}

@@

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files available on github
\url{https://github.com/kazewong/TurboPE/tree/main/src}.  We carry out these
runs using \textsc{parallel Bilby} (\textsc{pBilby}) \citep{Smith:2019ucc} to
distribute the computation over 400 Intel Skylake CPUs for each event. For the
specific settings chosen, the wall-time duration of each run was ${\sim}2$ h for
GW150914 and ${\sim}1$ day for GW170817. \corr{We have also benchmarked the
performance using a ROQ likelihood, which takes about 30 minutes for GW170817.
Without ROQ, it takes about a week to complete a run. This broadly agrees with
what is suggested in the literature \citep{Smith:2016qas}}

@@

% Additional traits for having differentiable waveforms
We are currently working on a number of improvements and extensions to our
current infrastructure. While the \textsc{IMRPhenomD} waveform approximant is a
reasonable start, it lacks some qualitative features that other state-of-the-art
models have, such as precession, subdominants moments of the radiation, tides,
or eccentricity. It also has a higher mismatch with reference numerical
relativity waveforms compared to more recent waveform models. Currently, we are
working on building differentiable implementations of \textsc{IMRPhenomXPHM}
\citep{Pratten:2020ceb}, a precessing successor to \textsc{IMRPhenomD} including
subdominant harmonics, as well as the numerical relativity surrogate waveforms,
including \textsc{NRSur7dq4} \citep{Varma:2018mmi}. Going forward, we expect the
use of autodifferentiation environments like \textsc{Jax} to become more
prevalent in the waveform development community, increasing the number of
differentiable waveform models available. This would not only be beneficial for
parameter estimation, but also for a number of other applications such as Fisher
matrix computations, template placement and calibrating waveforms to numerical
relativity
\citep{Coogan:2022qxs,Iacovelli:2022bbs,Iacovelli:2022mbg,Edwards:2023sak}.
\corr{Another family of waveform models that could benefit from this approach are
the effective-one-body (EOB) models \citep{Buonanno:1998gg,PhysRevD.89.061502}.
However, generating waveforms from the EOB family requires one to solve time-dependent
ODEs where the trunctation is dynamically determined at runtime.
JIT in \textsc{Jax} requires the computational graph of a function to be fixed
at compile time, which means the ODEs solving part of EOB is not compatible with
\textsc{Jax}. The models in the EOB family therefore may not benefit as much as the
other families.}
